quicR: An R Package for Real-Time Quaking Induced Conversion (RT-QuIC) Assays

### 2025-01-03 man/manuscript/manuscript.qmd
Moved the manuscript files to the man directory.

---
title: "quicR: An R Package for the Analysis of Real-Time Quaking Induced Conversion (RT-QuIC) Assays"
author:
    - name: 
        given: Gage
        family: Rowden
      id: GR
      orcid: 0000-0002-7517-0480
      email: rowde002@umn.edu
      affiliation:
          name: University of Minnesota
          city: Saint Paul
          state: MN
      corresponding: true
abstract: Real-time quaking induced conversion (RT-QuIC) has quickly become an emerging diagnostic tool for protein misfolding disorders such as Creutzfeldt-Jakob disease and Parkinson's disease. Given that the technology is still relatively new, academic and industry standards have yet to be established. 'quicR' was developed to fill this lack of standardization by providing functions for both data curation and analysis.
keywords: quicR, R package, RT-QuIC, prion, CJD, Parkinson's
bibliography:
    references.bib
bibliographystyle: 
    apa
---
```{r, echo=FALSE, message=FALSE}
library(quicR)
library(dplyr)
library(kableExtra)
library(readxl)
library(readr)

file <- "inst/extdata/input_files/test3.xlsx"
sample_file <- "inst/extdata/BMG_formatting/plate_layout.csv"
```

@@

## Development
The quicR package was developed to address the need for efficient data conversion and analysis of RT-QuIC data. The functions were designed with usability and reproducibility in mind, ensuring compatibility between multiple labs. Currently, the package accepts data exported from the MARS software as an Excel workbook.

@@

```{r, echo=FALSE, message=FALSE}
read_csv(sample_file, name_repair = "minimal") %>% kable()
```

@@

"BMG_format" includes the logical argument "write_file". If this argument is true, it will write a TXT file to the "save_path" as "save_name".

```{r}
BMG_format(sample_file, write_file = FALSE)[1:18]
```

@@

tabs <- convert_tables(tabs)
tabs %>% 
  head(15)
```

@@

df_[1:15, 1:6]
```

@@

```{r}
transpose_real(df_)[1:15, 1:13]
```

@@

df_norm[1:15, 1:12] %>%
  mutate_at(2:ncol(.), ~round(as.numeric(.), 2))
```

@@

### Time-to-threshold
Time-to-threshold is calculated using the "calculate_TtT" function.

### Maxpoint Ratio

### Maximum Slope

@@

data.frame("Sample IDs" = samples) %>%
  mutate(
    # Maxpoint Ratio
    MPR = calculate_MPR(df_norm, start_col = 3, data_is_norm = TRUE),
    # Max Slope
    MS = calculate_MS(df_norm, data_is_norm = TRUE),
    # Time to Threshold
    TtT = calculate_TtT(df_norm, threshold = 2, start_col = 3),
    # Rate of Amyloid Formation
    RAF = ifelse(TtT == hours, 0, 1 / TtT),
    # Crossed threshold?
    crossed = TtT != hours
  ) %>%
  head(15)
```

>>>>

---
title: "quicR: An R Package for Real-Time Quaking Induced Conversion (RT-QuIC) Assays"
author:
    - name: 
        given: Gage
        family: Rowden
      id: GR
      orcid: 0000-0002-7517-0480
      email: rowde002@umn.edu
      affiliation:
        name: University of Minnesota
        city: Saint Paul
        state: MN
      roles:
        Conceptualization
        Methodology
        Software
        Validation
        Formal analysis
        Investigation
        Resources
        Data curation
        Writing--original draft preparation
        Writing--review and editing
        Visualization
        Supervision
        Prject administration
      corresponding: true
abstract: Real-time quaking induced conversion (RT-QuIC) has quickly become an emerging diagnostic tool for protein misfolding disorders such as Creutzfeldt-Jakob disease and Parkinson's disease. Given that the technology is still relatively new, academic and industry standards have yet to be established. 'quicR' was developed to fill this lack of standardization by providing functions for both data curation and analysis.
keywords: quicR, R package, RT-QuIC, prion, CJD, Parkinson's
format: pdf
mainfont: Calibri
bibliography: references.bib
bibliographystyle: apa
---
```{r, echo=FALSE, message=FALSE}
library(quicR)
library(dplyr)
library(kableExtra)
library(readxl)
library(readr)

file <- "../../inst/extdata/input_files/test3.xlsx"
sample_file <- "../../inst/extdata/BMG_formatting/plate_layout.csv"
```

@@

# Development
The quicR package was developed to address the need for efficient data conversion and analysis of RT-QuIC data. The functions were designed with usability and reproducibility in mind, ensuring compatibility between multiple labs. Currently, the package accepts data exported from the MARS software as an Excel workbook.

@@

\begin{table}[ht]
\centering
\begin{tabular}{c|cccccccccccc}
   & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 \\ 
  \hline
  A & N & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
  B & N & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
  C & N & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
  D & N & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
  E & P & S12 & S13 & S14 & S15 & S16 & S17 & S18 & S19 & S20 & S21 & S22 \\ 
  F & P & S12 & S13 & S14 & S15 & S16 & S17 & S18 & S19 & S20 & S21 & S22 \\ 
  G & P & S12 & S13 & S14 & S15 & S16 & S17 & S18 & S19 & S20 & S21 & S22 \\ 
  H & P & S12 & S13 & S14 & S15 & S16 & S17 & S18 & S19 & S20 & S21 & S22 \\ 
   \hline
\end{tabular}
\end{table}

\newpage

@@

The function, "BMG_format", includes the logical argument "write_file". If true, it will create a TXT file. The path can be given to the "save_path" argument, and the file name can be supplied to the "save_name" argument.

```{r, message=FALSE}
BMG_format(sample_file, write_file = FALSE)[1:18] %>% 
  as.data.frame()
```

@@

tabs <- convert_tables(tabs)
tabs %>% head(15) %>% kable(row.names = FALSE)
```

@@

df_[1:15, 1:6] %>% kable(row.names = FALSE)
```

@@

```{r}
transpose_real(df_)[1:15, 1:13] %>% kable(row.names = FALSE)
```

@@

df_norm[1:15, 1:12] %>%
  mutate_at(2:ncol(.), ~round(as.numeric(.), 2)) %>%
  kable(row.names = FALSE)
```

@@

### Time-to-threshold
Time-to-threshold is calculated using the "calculate_TtT" function. The function must be supplied a threshold; default value is 2 (i.e. twice the background fluorescence if the data is normalized). A starting column should also be given as an integer; default value is 3. This is essentially asking how many columns of metadata are included before the fluorescence reads begin.

Time-to-threshold is calculated by iterating through each row and checking if a value is greater than the threshold. If the value is greater, the slope of the previous timepoint to the current timepoint is calculated, and the time intersection of the current read is returned.

### Maxpoint Ratio
Maxpoint ratios (MPR) are calculated by the "calculate_MPR" function. Data must be normalized in order to derive this metric. In a normalized data set, the MPR is simply the maximum value achieved during the run.

### Maximum Slope
The Maximum Slope (MS) is calculated by the "calculate_MS" function. The data can be either raw or normalized. The function iterates through each row using a rolling window which can be adjusted (default value is 3). Given the window size, a linear model is developed on those data points, and the slope of the model is recorded. The MS is simply the largest slope value recorded.

@@

data.frame("Sample IDs" = samples) %>%
  mutate(
    # Maxpoint Ratio
    MPR = calculate_MPR(df_norm, start_col = 3, data_is_norm = TRUE),
    # Max Slope
    MS = calculate_MS(df_norm, data_is_norm = TRUE),
    # Time to Threshold
    TtT = calculate_TtT(df_norm, threshold = 2, start_col = 3),
    # Rate of Amyloid Formation
    RAF = ifelse(TtT == hours, 0, 1 / TtT),
    # Crossed threshold?
    crossed = TtT != hours
  ) %>%
  head(15) %>%
  kable(row.names = FALSE)
```
